\chapter{Macroeconomic Crises}\label{chp:cris}
\hypertarget{crises}{}

\textbf{Tools:} Crisis triggers and indicators.

\textbf{Key Words:} Sovereign default; bank runs and panics; refinancing (rollover) risk; leverage; conditionality;
solvency and liquidity.

\textbf{Big Ideas:}
\vspace{-0.1in}
\begin{itemize}
\item Common triggers of macroeconomic crises are sovereign debt problems, financial fragility,
and fixed exchange rates.

\item Measures related to these triggers can help identify countries in trouble:
debt and deficits, financial weakness, exchange rate regime, and so on.

\item The goals of crisis prevention and crisis management are often at odds.
\end{itemize}
\rule{\textwidth}{1pt}

Economies periodically experience {\it crises\/}:
economic downturns that are not only larger than typical recessions
but qualitatively different.
The idea is now fresh in our minds,
but similar episodes have occurred throughout recorded history.
They're less common in modern, developed countries,
but they can happen anywhere.
Like snowflakes and business cycles, no two are exactly the same,
but they share some common features.


\section{Classic crisis triggers}

There are three classic triggers of macroeconomic crises:
sovereign debt, financial fragility, and a fixed exchange rate.

\textbf{Sovereign debt problems.}
If investors fear that a government may not repay its debt,
the market for debt collapses, often taking the economy with it.
In the old days, wars were the standard problem.
Wars are expensive, and if investors thought the expense was more
than the government was willing or able to bear,
they would stop buying the debt.
In modern times, governments spend money on many things besides wars,
but the possibility of default remains.
Argentina in 2002 and Greece today are recent examples.
These experiences remind us that sovereign debt need not be risk-free.

The central issue with government debts is sovereignty.
If a corporation defaults, the creditors take it to court
and claim the assets.
With governments, there's no such mechanism, and the process
is sloppier as a result.

\textbf{Financial fragility.}
We know from centuries of experience that when the
financial system freezes up,
economic activity slows down sharply.
We saw that in 2008,
but the same thing happened during the Bank of England Panic of 1825,
the Baring Crisis of 1890,
the US Panic of 1907,
Japan and Scandinavia in the 1990s,
and many other occasions.
It's a feature of even advanced financial systems that they
sometimes break.

In most cases, these financial problems follow from
poor investments (real estate is a common example),
which put the solvency of financial institutions in question.
The problems tend to snowball: Worries about the viability of one firm may lead others to
reduce their lending, leading to a cycle of retrenchment
that puts even sound firms in trouble.
The word ``panic'' is apt here and stems
from the imperfect information that investors have
when deciding where to put their money.


\textbf{Fixed exchange rates.} For whatever reason, fixed exchange rates periodically
must be defended in ways that undermine the economy.
Recent examples include the UK in 1992,
Mexico in 1994,
Korea in 1997,
and Argentina in 2000.

Many crises combine several of these elements.
The countries of the euro area face all three.
In Ireland and Spain, bailing out their banks
landed the governments in financial peril.
In other countries, bank positions in government debt
or in overpriced real estate put the banking systems in peril.
Finally, the common currency eliminates exchange rate changes
as a possible correction mechanism.


\section{Crisis indicators}

Crises are inherently hard to predict.
Why?
Think about cardiologists.
We understand that they can identify risk factors
(weight, high blood pressure) but cannot predict
the date of a heart attack with any precision.
Crises are worse.
Once people see a crisis on the way,
their actions tend to reinforce it:
they sell government debt,
withdraw funds from banks,
or shift their money to foreign currency.
But like a cardiologist, we can use what we know
to identify signs of trouble.


Analysts differ in the details, but most would include
the following in their ``checklist'' of crisis indicators:

%\begin{itemize}
\textbf{Government debt and deficits.}
The primary issue here is the quality of governance.
That aside, common rules of thumb include:
Worry if government
deficit is more than 5 percent of GDP or debt is more than 50 percent of GDP.
Adjust upward for developed countries, downward for developing countries
and for regional governments.
And watch out for hidden liabilities:  pensions, health care, bailouts, etc.
These are often much larger than official liabilities.

Fine points:  Worry further if debt is short-term and/or denominated in foreign currency.
Short-term debt subjects the government to refinancing (``rollover'') risk;
markets may demand better terms or refuse to refinance.
Foreign-denominated debt subjects government to risk
if currency falls in value, making the debt larger in local terms.

\textbf{Banking/financial system.}
This isn't something we've discussed,
but  analysts track leverage, duration mismatch, exposure concentration,
risk-management processes,
and nonperforming loans.
The challenge is measuring them accurately from reported information.
Some of the most troubling situations come with low-quality data.


\textbf{Exchange rate and reserves.}
Rule of thumb:  Worry if the exchange rate is fixed, or close to it,
and the currency is significantly overvalued in PPP terms
(Big Macs cost 30 percent more than in other currencies;
the real exchange rate has risen more than 30 percent in the past 2-5 years).
Worry more if foreign-exchange reserves
are low or have fallen significantly.

\textbf{Political situation.}
%Rule of thumb:  Worry if the political system seems unable or unwilling
%to deal with problems that could lead to crises.
Crises are often more political than economic.
Countries with effective governments suffer fewer crises and
deal with those that occur more effectively.
Analysts therefore look for signs that the political
system is unable or unwilling to deal with problems that might turn into crises.
Zimbabwe's hyperinflation and Argentina's debt crisis are good examples.
Or the Weimar Republic in 1920s Germany.
%\end{itemize}

All of these things generate more concern
in countries with weak institutions.
It's not an accident that Greece is in worse trouble
than France or Germany.
%Analysts often find that local politics are more important
%than the numbers to the outcome.


\section{Crisis responses}

What should a government do when faced with a crisis?
It depends on the trigger.
Standard advice includes:

\textbf{Sovereign debt crises.}
If the problem is that the government is borrowing too much,
the answer is to stop doing it --- run primary surpluses until the debt is manageable.
Default is also an option, and saves money in the short term,
but probably raises borrowing costs in the future.
And if you go through a default, it's helpful to resolve it as quickly as possible.
IMF support is often used to cushion the blow: Contingent on progress with the deficit,
the IMF lends the government money on more attractive terms
than the market would provide.
This ``conditionality,'' as it's called,
helps reduce moral hazard (you get the money only if you behave)
and provides cover for local politicians (the IMF made us do this).
Such conditional lending can be critical in a crisis,
when high borrowing rates exacerbate the government's debt problems.
(See:  debt dynamics.)


\textbf{Financial crises.}
If the financial system is fundamentally sound (solvent) but illiquid,
the longstanding advice is for the central bank to lend aggressively.
The standard quote comes from Walter Bagehot,
a 19th-century businessman and journalist:
``To avert panic, central banks should lend early and freely,
to solvent firms, against good collateral, and at high rates.''
%Put simply,
%the goals of crisis prevention and crisis management
%are often at odds.


If the financial system is insolvent,
it's important to get it recapitalized and operating again.
This advice comes with more than a little irony,
as governments sometimes find themselves bailing out
precisely those banks that triggered the crisis.
The trick is to do it in ways that inflict some pain
on the bank's management and creditors
(incentives for the future) and don't bankrupt the government.
These things happen fast, so it's hard to get everything right.


\textbf{Fixed exchange rates.} Let them float.
A more controversial approach is to impose capital controls to inhibit the
response of capital markets to a possible drop in the exchange rate.
(See:  trilemma.)
Capital controls are a dangerous tool,
because the fear of future capital controls
can generate a crisis on its own,
as investors rush to get their money out of the country.
Capital controls on inflows may, for that reason, be more attractive
than controls on outflows.


\section*{Executive summary}

\setlength{\leftmargini}{.5\oldleftmargini}
\begin{enumerate}\itemsep=0.0in
\item The classic crisis triggers are
(i)~government debt and deficits,
(ii)~a fragile financial system,
and (iii)~a fixed exchange-rate system.

\item Crises are hard to predict,
but we nevertheless have useful indicators connected to each
of the triggers.

\item Politics and institutions are central.
\end{enumerate}
\setlength{\leftmargini}{\oldleftmargini}

%\end{document}

\section*{Review questions}

\setlength{\leftmargini}{.5\oldleftmargini}
\begin{enumerate}
\item Risk and opportunity in Ghana (May 2012).
You have been asked to prepare a risk assessment for
the West African country of Ghana.
Ghana is a former British colony that has been growing rapidly
in recent years after a period of unusually stable politics.
The Economist Intelligence Unit refers to it as a ``robust democracy.''
The World Economic Forum ranked Ghana 114th (of 133)
in their Global Competitiveness Report.
They continue:
``The country continues to display strong public institutions and
governance indicators,
particularly in regional comparison.''

The EIU's Country Risk Report adds:
\begin{itemize}
\item The December 2012 elections are expected to be close.
The president, John Atta Mills, came to power promising accountability
and transparency, but  has struggled to maintain party unity
while evidence emerges of financial impropriety of some government ministers.
\item The victor faces a challenging policy environment, particularly
the fiscal situation.
\item Expectations among the population are high as production
starts at the offshore Jubilee oil field.
\item The government's decision to allow use of 70\% of future
oil revenue as collateral for borrowing is a cause for concern
if the revenue is not managed properly.
\item The Bank of Ghana (the central bank) faces the twin goals of
containing inflation and fostering growth.
\item The currency  ---  the cedi --- floats with occasional heavy intervention.
\end{itemize}
%
Your mission is to assess the risks to Ghana  using
the information in the table, as well as
your own good judgement and analytical skills.

\begin{enumerate}
\item  You decide to start with a fiscal assessment.
What trend do you see in government revenues and expenses?

\item You notice that neither the primary deficit nor
interest expenses are reported separately.
How would you estimate them from the numbers in the table?
What are their values for 2011?

\item Using what you know about government debt dynamics,
compute the ratio of government debt to GDP for 2011.
What factors contribute the most to the change from 2010?

\item Overall, how would you assess the risks to Ghana's economy over
the next couple of years?
\end{enumerate}

{\small
\begin{tabular}{lrrrrr}
\toprule
        & 2008 & 2009 & 2010 & 2011 & 2012 \\
\midrule
GDP growth (\%) & 8.4 & 4.0 & 7.7 & 13.6 & 7.4 \\
Inflation (\%)  & 18.1 & 16.0 & 8.6 & 8.6 & 8.5 \\
Interest rate (\%) & 20.8 & 28.8 & 22.7 & 20.5 & 20.6  \\
Govt revenue (\% of GDP)  & 16.0 & 16.5 & 19.1 & 23.4 & 22.2 \\
Govt spending (\% of GDP) & 24.5 & 22.3 & 25.5 & 27.6 & 27.7 \\
Govt budget balance (\% of GDP) & --8.5 & --5.8 & --6.5& --4.2 & --5.5\\
Govt debt (\% of GDP) & 30.6 & 33.3 & 33.9 & {\bf } & {\bf } \\ %& 36.8 & 41.6\\
Real exchange rate (index) & 81.7 & 76.3 & 81.8 & 78.1 & 74.5\\
FX reserves (USD billions) & 1.8 & 2.9 & 4.3 & 4.4 & 4.8 \\
\bottomrule
\end{tabular}
}

\medskip
Data from EIU CountryData.
The government budget balance is a surplus if positive, deficit if negative.
The real exchange rate is the price of goods in Ghana relative to the rest
of the world;
the larger the number, the more expensive goods are in Ghana.
The numbers for 2011 and 2012 are estimates.

Answer.
\begin{enumerate}
\item Trends include:
(i)~revenues and spending both rising,
(ii)~spending still ahead of revenue (there's a deficit),
and (as a direct result)
(iii)~ratio of debt to GDP rising a little
(more on that to come).

\item This is a tricky one.  
Remember that interest payments in year $t$ are $ i_t B_{t-1}/Y_t$.
We get what we want from:
\begin{eqnarray*}
    i_t B_{t-1}/ Y_{t} &=& i_t (B_{t-1}/ Y_{t-1}) (Y_{t-1}/ Y_{t}) \\
           &\approx&  i_t (B_{t-1}/Y_{t-1}) /(1+g_t + \pi_t) .
\end{eqnarray*}
That gives us interest payments in 2011 of 5.7\% of GDP and
a primary deficit of $-1.5$\% (that is, a surplus).

\item The key relation is this one:
\begin{eqnarray*}
    \Delta ({B_{t}}/{Y_{t}})
            &=&
                (i_t-\pi_t) ({B_{t-1}}/{Y_{t-1}})
                - g_t ({B_{t-1}}/{Y_{t-1}})
             +    ({D_{t}}/{Y_{t}})  .
\end{eqnarray*}
We refer to the components on the right as A, B, and C.
Doing the calculations gives us

\begin{center}
\begin{tabular}{lrr}
\toprule
        &  2010 & 2011   \\
\midrule
Interest payments  &  &  4.0  \\
Component A (interest)  &  &  4.0  \\
Component B (growth)            &   & --4.6  \\
Component C (primary deficit)   &   & --1.5  \\
Total change in $B/Y$       &       & --2.1   \\
Public debt (\% of GDP)     &  33.9 & 31.8   \\
\bottomrule
\end{tabular}
\end{center}

Over this period, the ratio of debt to GDP fell by 2.1\%.
The components contributed:
interest +4.0, growth --4.6, and the primary deficit --1.5.
Note especially the growth term, the result of unusually high GDP growth in 2011.

\item This is a call to look at the checklist of crisis indicators:
\begin{itemize}
\item Government debt and deficits.
We have deficits, but there's not
much sign yet of a growth debt to GDP ratio.
One future concern might be the possibility of borrowing now against future oil
revenue.  Will any debts incurred be spent wisely?
Will the oil revenue show up?
\item Banking system.  No information provided.
\item Exchange rate and reserves.  Reserves are modest,
but with the exchange rate floating there shouldn't be much
concern about that.
\item Politics.  Always an issue,
especially with a contentious election coming
and the promise of money from oil revenue.
It's an odd fact but a true one that revenue
from natural resources is more likely to cause problems than solve them.
\end{itemize}
\end{enumerate}

\end{enumerate}
\setlength{\leftmargini}{\oldleftmargini}


\section*{If you're looking for more}

You can find similar analyses in many places.
One of our favorites is the 
Economist Intelligence Unit's Country Risk Reports.
Another is the IMF's
\href{http://www.imf.org/external/np/exr/facts/vul.htm}
{Vulnerability Indicators}.

There's no end of good descriptions of crises.
On the most recent crisis, Ben Bernanke's
\href{http://www.federalreserve.gov/newsevents/testimony/bernanke20100902a.htm}
{testimony} to the crisis commission is a good overview
from the perspective of the US
(search:  ``bernanke testimony crisis causes'').
Michael Lewis's Vanity Fair pieces are works of art
(search ``michael lewis vanity fair'').
Among the many books, we recommend
%
\begin{itemize}
\item Robert Bruner and Sean Carr,
\href{http://www.amazon.com/Panic-1907-Lessons-Learned-Markets/dp/0470452587/}
{\it The Panic of 1907\/}.
Good read, and short; you'll think it's about 2007.

\item Carmen Reinhart and Kenneth Rogoff,
\href{http://www.amazon.com/This-Time-Different-Centuries-ebook/dp/B004EYT932/}
{\it This Time is Different\/}.
The recent bestseller, covering 800 years and the whole world.

\item David Wessel,
\href{http://www.amazon.com/Fed-We-Trust-Bernankes-Great/dp/0307459683/}
{\it In Fed We Trust\/}.
Terrific book from the \emph{Wall Street Journal} writer.

%\item Paul Blustein,
%    {\it And the money rolled in (and out).\/}
%    Wonderful review of Argentina's 1999-2001 crisis.
%    He has another one, {\it The chastening\/}, about the Asian crisis
%    and the IMF.
%\item Andres Oppenheimer, {\it Bordering on chaos: Mexico's roller-coaster journey %toward prosperity.\/}
%    About Mexico in running up to the election and
%    subsequent crisis of 1994.
\end{itemize}


